Impact of grids and dynamical cores in CESM2.2 on the surface mass balance of the Greenland Ice Sheet

### 2022-05-07 arcticgrids.tex
mods

Figure~\ref{fig:tseries} shows time series of annually integrated precipitation and snow/ice melt over the GrIS for the various different grids and dycores, and RACMO in black. The 1979-1998 climatological mean values from Table~\ref{tbl:table3} are shown as circles on the right side of the panels. The uniform $1^{\circ}-2^{\circ}$ grids have positive precipitation biases, whereas the VR grids have the smallest biases, with precipitation comparable to RACMO. The $\tt f19$ and $\tt f09$ grids perform similarly, with +110 Gt/yr bias, whereas $\tt ne30pg3$ is biased by about +165 Gt/yr and $\tt ne30pg2$ by +230 Gt/yr. The larger biases on the uniform-resolution SE grids relative to the FV grids are consistent with the coarser GrIS resolution on the SE grids (Figure~\ref{fig:grisdx}).

@@

Lower troposphere temperature is not a strict proxy for melting; e.g., it may not capture microclimate effects as a result of a better representation of the low-elevation ablation zones. Positive degree-days \cite<PDD;>{B1984Z}, which accumulate the near-surface temperature in $^{\circ}C$ for days with temperature above freezing, are a more accurate proxy.
PDD is nonlinear in mean monthly temperature \cite{R1991P}.  We compute it from monthly mean 2-meter temperature using the method of \citeA{CG2005JG}, assuming a fixed monthly mean standard deviation of 3$^{\circ}$C and a degree-day factor of 5~mm~d$^{-1}$ $^{\circ}$C$^{-1}$.

@@

The bottom panel of Figure~\ref{fig:prect} presents the biases in the combined ice/snow melt as map plots. These plots show that the largest melt biases are on the southeast and northwest coasts, where large coarse-grid cells overlap with the ocean. One possibility is that these problematic grid cells are situated at lower elevations than the true ice sheet surface, leading to a warm bias and too much melt. Figure~\ref{fig:ztrans} shows the representation of the ice sheet surface along two transects on the different grids, compared to the high-resolution dataset used to generate CAM topographic boundary conditions \cite{GMTED2010,gmdd-8-4623-2015}. The two transects are shown in Figure~\ref{fig:prect}: the east-west ``K-transect'' in southwest Greenland and a transect extending from the central dome down to the  Kangerlussuaq glacier on the southeast coast. The $1^{\circ}-2^{\circ}$ grids are noticeably coarse, with only a handful of grid cells populating the transect. The $\tt f09$ grid is a bit of an exception, since the grid cells become narrow in the meridional direction at high latitudes, and a larger number of grid cells populate an east-west transect. The VR grids are more skillful at reproducing the steep margins of the ice sheet, capturing the parabolic shape of the GrIS margins.

@@

The transects in Figure~\ref{fig:ztrans} show that the ice sheet surface on the coarse grids is not systematically lower than the true surface in ablation zones. Rather, the smoothing and flattening of the raw topography, necessary to prevent the model from exciting grid-scale numerical modes, causes the lower-elevation ablation zones to extend beyond the true ice sheet margin, where they lie above the actual ice surface. The $\tt f19$ grid has both the smoothest topography and the flattest ice sheet since its dynamics are coarsest ($\tt f09$, $\tt ne30pg2$ and $\tt ne30pg3$ use identical smoothing). This suggests that coarser models will tend to elevate the ablation zones and thereby depress melt rates.

@@

Figure~\ref{fig:bias} shows mean melt bias, relative to both RACMO datasets, conditionally sampled based on grid cell ice fraction in the GrIS region. Errors are computed using the common ice mask approach described in section~\ref{sec:SMB}, and so the grid cell ice fraction are taken from the common masks. The figure shows that coarser grids generally have two peaks in ice fraction space; a bump in positive melting errors in the 0-20$\%$ range, and another in the fully ice-covered cells. Also shown are the $\pm 1$~standard deviation of the biases for each bin. They indicate that the biases in 0-20$\%$ bins are mostly contained in the positive bias region (a fractional bias greater than 0), whereas the fully-covered ice cells have a wider distribution, with many grid cells also containing negative melting biases. The excessive melting in the 0-20$\%$ ice fraction bins supports our hypothesis that the prevalence of mixed-grid cells in the ablation zone on coarse grids is responsible for their large melt bias.

({\color{blue}{Any idea why the errors are smaller in the cells with intermediate ice fraction?  I wouldn't have expected this.}})
{\color{green}{Rene - is the melt map consistent with the bin figure? Smallest errors are interior points which correspond to glacier fraction bin of 1, which the largest errors.}}
{\color{blue}{ARH - I can check. The bin figure is fractional change so its a different metric, which could explain the apparent inconsistency.}}

@@

Synoptic storms are tracked using TempestExtremes atmospheric feature detection software \cite{UETAL2021}. As the $\tt Arctic$ grid contains $1/4^{\circ}$ refinement north of about $45^{\circ}$ latitude, the storm tracker is applied to this region for the $\tt Arctic$ and $\tt ne30pg3$ runs to identify differences in storm characteristics due to horizontal resolution. The composite mean precipitation maps are similar between the two grids, and exhibit the iconic comma structure of synoptic cyclones (not shown). {\color{purple}{Marcus - no need to mention if not shown.}}

>>>>

Figure~\ref{fig:tseries} shows time series of annually integrated precipitation and snow/ice melt over the GrIS for the various different grids and dycores, and RACMO in black. The 1979-1998 climatological mean values from Table~\ref{tbl:table3} are shown as circles on the right side of the panels. The uniform $1^{\circ}-2^{\circ}$ grids have positive precipitation biases, whereas the VR grids have the smallest biases, with precipitation comparable to RACMO. The $\tt f19$ and $\tt f09$ grids perform similarly, with +110 Gt/yr bias, whereas $\tt ne30pg3$ is biased by about +165 Gt/yr and $\tt ne30pg2$ by +230 Gt/yr. The larger biases on the uniform-resolution SE grids relative to the FV grids are consistent with the coarser GrIS resolution in the SE grids (Figure~\ref{fig:grisdx}).

@@

Lower troposphere temperature is not a strict proxy for melting; e.g., it may not capture microclimate effects as a result of a better representation of the low-elevation ablation zones. The Positive degree-day temperature based melt index \cite<PDD;>{B1984Z}, which accumulate the near-surface temperature in $^{\circ}C$ for days with temperature above freezing, is a more accurate proxy for melting. PDD is nonlinear in mean monthly temperature \cite{R1991P}.  We compute PDD from monthly mean 2-meter temperature using the method of \citeA{CG2005JG}, assuming a fixed monthly mean standard deviation of 3$^{\circ}$C and a degree-day factor of 5~mm~d$^{-1}$ $^{\circ}$C$^{-1}$.

@@

The bottom panel of Figure~\ref{fig:prect} presents the biases in the combined ice/snow melt as map plots. These plots show that the largest melt biases are on the southeast and northwest coasts, where large coarse-grid cells overlap with the ocean. One possibility is that these problematic grid cells are situated at lower elevations than the true ice sheet surface, leading to a warm bias and too much melt. Figure~\ref{fig:ztrans} shows the representation of the ice sheet surface along two transects on the different grids, compared to the high-resolution dataset used to generate CAM topographic boundary conditions \cite{GMTED2010,gmdd-8-4623-2015}. The two transects are shown in Figure~\ref{fig:prect}: the east-west ``K-transect'' in southwest Greenland and a transect extending from the central dome down to the  Kangerlussuaq glacier on the southeast coast. The $1^{\circ}-2^{\circ}$ grids are noticeably coarse, with only a handful of grid cells populating the transect. The $\tt f09$ grid is a bit of an exception since the grid cells narrow in the zonal direction at high latitudes, and a larger number of grid cells populate an east-west transect. The VR grids are more skillful at reproducing the steep margins of the ice sheet, capturing the parabolic shape of the GrIS margin.

@@

The transects in Figure~\ref{fig:ztrans} show that the ice sheet surface on the coarse grids is not systematically lower than the true surface in ablation zones. Rather, the smoothing and flattening of the raw topography, necessary to prevent the model from exciting grid-scale numerical modes, causes the lower-elevation ablation zones to extend beyond the true ice sheet margin, where they lie above the actual ice surface. The $\tt f19$ grid has both the smoothest topography and the flattest ice sheet since its dynamics are coarsest, whereas the $\tt f09$, $\tt ne30pg2$ and $\tt ne30pg3$ grids use identical smoothing. This suggests that coarser models will tend to elevate the ablation zones and thereby depress melt rates, which is the opposite of what occurs in the coarse grid simulations.

@@

Figure~\ref{fig:bias} shows the mean melt bias, relative to both RACMO datasets, conditionally sampled based on grid cell ice fraction in the GrIS region. Errors are computed after mapping to the common ice masks using different methods, described in section~\ref{sec:SMB}. The grid cell ice fractions therefore pertain to ice fractions on the low-resolution common ice masks. Also shown are the $\pm 1$~standard deviation of the biases for each bin. The figure shows that coarser grids can be characterized by a monotonic increase in melt bias as the grid cell ice fraction decreases. The VR grids have the smallest melt biases for small grid cell ice fractions (smaller than 50\%), the uniform SE grids and $\tt f19$ have the largest melt biases and the $\tt f09$ grid melt biases lie between these two cases. Figure~\ref{fig:bias} generally supports our hypothesis that the prevalence of mixed-grid cells in the ablation zone in coarse grids is responsible for their large melt bias.

@@

Synoptic storms are tracked using TempestExtremes atmospheric feature detection software \cite{UETAL2021}. As the $\tt Arctic$ grid contains $1/4^{\circ}$ refinement north of about $45^{\circ}$ latitude, the storm tracker is applied to this region for the $\tt Arctic$ and $\tt ne30pg3$ runs to identify differences in storm characteristics due to horizontal resolution.
